---
title: "scratch_analysis"
author: "Holden Jones"
date: "2025-07-21"
output: html_document
---




## use neg binomial most likely

full Poisson glm for abundance
```{r}
# # full abundance model with all predictors
# glm_full_abundance <- glm(n ~ Tipo + veg_scaled + CC_scaled + forest_scaled + 
#                        water_scaled + elevation_scaled, 
#                        data = site_data,
#                        family = poisson)
# ```
# 
# use dredge to run model selection - limit to max of three predictors in a model
# ```{r}
# # make sure na.action is set globally for dredge
# options(na.action = "na.fail")  
# 
# # run dredge to get all possible model combinations
# abundance_model_set <- dredge(glm_full_abundance, m.lim = c(1,3))
# 
# # view top models by AIC
# head(abundance_model_set)
# 
# # get the best model (lowest AICc)
# best_abundance_model <- get.models(abundance_model_set, subset = 1)[[1]]
# 
# # summary of the best model
# summary(best_abundance_model)
# 
# # assess multicollinearity
# vif(glm_full_abundance)
# vif(best_abundance_model)
# 
# # assess how best model fits with DHARMa package
# simulationOutput <- simulateResiduals(fittedModel = best_abundance_model)
# plot(simulationOutput)
# 
# testDispersion(simulationOutput)
# testZeroInflation(simulationOutput)
```






### negative binomial
Poisson looked good for richness - but still check out negative binomial to compare

full negative binomial glm.nb for richness
```{r}
glm_full_richness <- glm.nb(no_species ~ Tipo + veg_scaled + CC_scaled + forest_scaled + 
                       water_scaled + elevation_scaled, 
                       data = site_data)
#Warning in theta.ml(Y, mu, sum(w), w, limit = control$maxit, trace = control$trace >  :iteration limit reached
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")  

# run dredge to get all possible model combinations - test restricting or not
richness_model_set <- dredge(glm_full_richness, m.lim = c(1,3))

# view top models by AIC
head(richness_model_set)

# get the best model (lowest AICc)
best_richness_model <- get.models(richness_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_richness_model)

# assess multicollinearity
vif(glm_full_richness)
# vif(best_richness_model) # contains fewer than two terms

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_richness_model)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)

# assess how full model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_full_richness)
plot(simulationOutput) # combined adjusted quantile test significant

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```






## figures





## publication level visualization - working

```{r}
# # plot Shannon diversity between paired sites using palet, my_theme
# plot_paired_shannon <- ggplot(paired_data, 
#                                      aes(x = Tipo, y = Shannon_Index, fill = Tipo)) +
#   geom_boxplot() +
#   scale_fill_manual(values = pal) + 
#   scale_x_discrete(labels = c(paired_site_labels)) +
#   my_theme() + 
#   labs(x = "Cacao type",
#        y = "Shannon Index")
# plot_paired_shannon
# #ggsave("output/plot_paired_shannon.png", width = 10, height = 10)
```







## publication level visualization - working

```{r}
# plot Simpson diversity between paired sites using palet, my_theme
# plot_paired_simpson <- ggplot(paired_data, 
#                                      aes(x = Tipo, y = Simpson_Index, fill = Tipo)) +
#   geom_boxplot() +
#   scale_fill_manual(values = pal) + 
#   scale_x_discrete(labels = c(paired_site_labels)) +
#   my_theme() +
#   labs(x = "Cacao type",
#        y = "Simpson Index") + 
#   annotate("text", x = 1.5, y = max(paired_data$Simpson_Index, na.rm = TRUE) * 1.05, 
#            label = "ns", size = 6)
# 
# plot_paired_simpson
#ggsave("output/plot_paired_simpson.png", width = 10, height = 10)


## UNSURE IF THIS DOES ANYTHING
# plot_paired_simpson <- plot_paired_simpson +
#   theme(
#     axis.title.y = element_text(
#       margin = margin(t = 200),  # increase this to nudge downward
#       size = 22,
#       color = "gray25"
#     )
#   )
# plot_paired_simpson
```






## publication level visualization - working

```{r}
# plot abundance between paired sites using palet, my_theme
plot_paired_abundance <- ggplot(paired_data, 
                                     aes(x = Tipo, y = n, fill = Tipo)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) + # order needs to match factor levels above
  scale_x_discrete(labels = c(paired_site_labels)) +
  my_theme() + 
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(y = "Abundance")



plot_paired_abundance
#ggsave("output/plot_paired_abundance.png", width = 10, height = 10)
```
