---
title: "pub_cacao_analysis"
author: "Holden Jones"
date: "2025-07-11"
output: html_document
---

Statistical tests and figures for cacao ranas publication
- using processed dataframes created in pub_cacao_process script


#1. Load

```{r}
library(tidyverse)
library(GGally)
library(car)
library(DHARMa)
library(ggResidpanel)
library(ggeffects)
library(MuMIn)
library(MASS) # note that this produces masking issues for select()
```

```{r}
site_data <- read_csv("data/processed_data/site_data_processed.csv")
```


#2. Paired sun and shade comparisons

Most of the cacao sites in the study are in adjacent shade and sun pairs
- can use paired t-tests to control for differences in location
- compare response variables (abundance, richness, diversity)
- note that one shade site was not paired, remove this site from this analysis


##a. create paired_data

```{r}
# retain only paired shade and sun cacao sites
paired_data <- site_data %>%
  filter(Tipo == 'N' | Tipo == 'C',
         !Sitio == 'VC-N')
```


##b. abundance

paired t-test
```{r}
# restructure paired_data to wide format for abundance
paired_abundance <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'n')) %>%
  pivot_wider(names_from = Tipo, values_from = n)

# run paired t-test
t.test(paired_abundance$N, paired_abundance$C, paired = TRUE)
```

visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = n, fill = Tipo)) +
  geom_boxplot()
```


##c. richness

paired t-test
```{r}
# restructure paired_data to wide format for richness
paired_richness <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'no_species')) %>%
  pivot_wider(names_from = Tipo, values_from = no_species)

# run paired t-test
t.test(paired_richness$N, paired_richness$C, paired = TRUE)
```

visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = no_species, fill = Tipo)) +
  geom_boxplot()
```


##d. diversity indices

Shannon diversity paired t-test
```{r}
# restructure paired_data to wide format for Shannon diversity
paired_shannon <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'Shannon_Index')) %>%
  pivot_wider(names_from = Tipo, values_from = Shannon_Index)

# run paired t-test
t.test(paired_shannon$N, paired_shannon$C, paired = TRUE)
```

Shannon diversity visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = Shannon_Index, fill = Tipo)) +
  geom_boxplot()
```

Simpson diversity paired t-test
```{r}
# restructure paired_data to wide format for Simpson diversity
paired_simpson <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'Simpson_Index')) %>%
  pivot_wider(names_from = Tipo, values_from = Simpson_Index)

# run paired t-test
t.test(paired_simpson$N, paired_simpson$C, paired = TRUE)
```

Simpson diversity visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = Simpson_Index, fill = Tipo)) +
  geom_boxplot()
```

inverse Simpson diversity paired t-test
```{r}
# restructure paired_data to wide format for inverse Simpson diversity
paired_inv_simpson <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'Inv_Simpson_Index')) %>%
  pivot_wider(names_from = Tipo, values_from = Inv_Simpson_Index)

# run paired t-test
t.test(paired_inv_simpson$N, paired_inv_simpson$C, paired = TRUE)
```

Shannon diversity visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = Inv_Simpson_Index, fill = Tipo)) +
  geom_boxplot()
```


#3. GLMs for each predictor variable

GLMs to assess the influence of local and landscape variables on response vars
- include all site types, site type as fixed effect

response vars
- abundance, no_species, Shannon, Simpson, Inv_Simpson

predictor vars
- tipo, scaled: veg, hoja, CC, forest, water, elevation


##a. assess distribution of response variables

histogram of each response var
- will assume Poisson distribution first b/c count data
- assess overdispersion and model fit during model building and adjust as needed
```{r}
hist(site_data$n) # assume Poisson
hist(site_data$no_species) # assume Poisson
hist(site_data$Shannon_Index) # approximately normal
hist(site_data$Simpson_Index) # approximately normal
hist(site_data$Inv_Simpson_Index) # perhaps lognormal?
```


##b. assess correlation of predictor variables

look at correlation matrix - high collinearity if correlation > 0.7
```{r}
ggpairs(site_data, columns = c(18,19,20,21,22,23))
```

canopy cover and hoja are highly correlated (0.873!) - only use canopy cover


##c. plot variable relationships with abundance

```{r}
ggplot(site_data, aes(x = Tipo, y = n)) +
  geom_boxplot()

ggplot(site_data, aes(x = veg_scaled, y = n)) +
  geom_point()

ggplot(site_data, aes(x = CC_scaled, y = n)) +
  geom_point()

ggplot(site_data, aes(x = forest_scaled, y = n)) +
  geom_point()

ggplot(site_data, aes(x = water_scaled, y = n)) +
  geom_point()

ggplot(site_data, aes(x = elevation_scaled, y = n)) +
  geom_point()
```


## d. GLM for abundance with all predictors

Poisson abundance model was overdispersed - use negative binomial instead

full negative binomial glm.nb for abundance
```{r}
glm_full_abundance <- glm.nb(n ~ Tipo + veg_scaled + CC_scaled + forest_scaled + 
                       water_scaled + elevation_scaled, 
                       data = site_data)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")  

# run dredge to get all possible model combinations - test restricting or not
abundance_model_set <- dredge(glm_full_abundance, m.lim = c(1,3))

# view top models by AIC
head(abundance_model_set)

# get the best model (lowest AICc)
best_abundance_model <- get.models(abundance_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_abundance_model)

# assess multicollinearity
vif(glm_full_abundance)
# vif(best_abundance_model) # contains fewer than two terms

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_abundance_model)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)

# assess how full model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_full_abundance)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

negative binomial abundance model fits well 
- best model contains only CC_scaled - although this is not sig


##e. GLM for richness with all predictors

same process as above except using richness as response variable here

Poisson seems okay for richness - best model only CC, not sig

full Poisson glm for richness
```{r}
# full richness model with all predictors
glm_full_richness <- glm(no_species ~ Tipo + veg_scaled + CC_scaled + forest_scaled +
                       water_scaled + elevation_scaled,
                       data = site_data,
                       family = poisson)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")

# run dredge to get all possible model combinations
richness_model_set <- dredge(glm_full_richness, m.lim = c(1,3))

# view top models by AIC
head(richness_model_set)

# get the best model (lowest AICc)
best_richness_model <- get.models(richness_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_richness_model)

# assess multicollinearity
vif(glm_full_richness)
# vif(best_richness_model) model contains fewer than two terms

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_richness_model)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput) # how do we interpret this?
```

Poisson looked good for richness - still check out negative binomial to compare

full negative binomial glm.nb for richness
```{r}
glm_full_richness <- glm.nb(n ~ Tipo + veg_scaled + CC_scaled + forest_scaled + 
                       water_scaled + elevation_scaled, 
                       data = site_data)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")  

# run dredge to get all possible model combinations - test restricting or not
richness_model_set <- dredge(glm_full_richness, m.lim = c(1,3))

# view top models by AIC
head(richness_model_set)

# get the best model (lowest AICc)
best_richness_model <- get.models(richness_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_richness_model)

# assess multicollinearity
vif(glm_full_richness)
# vif(best_richness_model) # contains fewer than two terms

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_richness_model)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)

# assess how full model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_full_richness)
plot(simulationOutput) # combined adjusted quantile test significant

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

##f. GLM for Shannon diversity with all predictors

elevation sig negative effect on shannon diversity. interesting
combined adjusted quantile test significant

full glm for shannon with normal dist.
- Shannon isn't count data! so can't use Poisson? non-integer values?
```{r}
# full shannon model with all predictors
glm_full_shannon <- glm(Shannon_Index ~ Tipo + veg_scaled + CC_scaled + forest_scaled +
                       water_scaled + elevation_scaled,
                       data = site_data,
                       family = gaussian)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")

# run dredge to get all possible model combinations
shannon_model_set <- dredge(glm_full_shannon, m.lim = c(1,3))

# view top models by AIC
head(shannon_model_set)

# get the best model (lowest AICc)
best_shannon_model <- get.models(shannon_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_shannon_model) # elevation sig negative effect on shannon

# assess multicollinearity
vif(glm_full_shannon)
vif(best_shannon_model) # inflation not an issue in best model

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_shannon_model)
plot(simulationOutput) # combined adjusted quantile test significant

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```


##g. GLM for Simpson diversity with all predictors

elevation sig negative effect on simpson diversity 
tipo has sig effect on simpson diversity - sig higher in bosque than N,C

full glm for simpson with normal dist.
- simpson isn't count data! so can't use Poisson? non-integer values?
```{r}
# full simpson model with all predictors
glm_full_simpson <- glm(Simpson_Index ~ Tipo + veg_scaled + CC_scaled + forest_scaled +
                       water_scaled + elevation_scaled,
                       data = site_data,
                       family = gaussian)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")

# run dredge to get all possible model combinations
simpson_model_set <- dredge(glm_full_simpson, m.lim = c(1,3))

# view top models by AIC
head(simpson_model_set)

# get the best model (lowest AICc)
best_simpson_model <- get.models(simpson_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_simpson_model) # elevation and tipo significant effect

# assess multicollinearity
vif(glm_full_simpson)
vif(best_simpson_model) # inflation not an issue in best model

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_simpson_model)
plot(simulationOutput) # combined adjusted quantile test not significant

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```
















