---
title: "pub_cacao_analysis"
author: "Holden Jones"
date: "2025-07-11"
output: html_document
---

Statistical tests and figures for cacao ranas publication
- using processed dataframes created in pub_cacao_process script


#1. Load

```{r}
library(tidyverse)
library(RColorBrewer)
library(GGally)
library(car)
library(ggpubr)
library(DHARMa)
library(ggResidpanel)
library(ggeffects)
library(MuMIn)
library(MASS) # note that this produces masking issues for select()
library(multcomp)
```

```{r}
site_data <- read_csv("data/processed_data/site_data_processed.csv")

# reorder factor levels for plotting
site_data$Tipo <- factor(site_data$Tipo,
                         c("C", "N", "V", "B"))
```

define a consistent palet, labels, and theme to be used in figures
```{r}
# palet for comparisons of all 4 site types
pal <- brewer.pal(4, "BrBG")

# needs to match factor levels above
pal <- c("#DFC27D", "#A6611A", "#80CDC1", "#018571")

# set a theme for ggplots
my_theme <- function() {
  theme_minimal() + 
   theme(legend.position = "none",
        plot.background = element_rect("white"),
        panel.background = element_rect("white"),
        panel.grid = element_line("grey90"),
        axis.line = element_line("gray25"),
        axis.text = element_text(size = 22, color = "gray25"),
        axis.title = element_text(size = 26, color = "gray25"),
        legend.text = element_text(size = 22))
}
```


#2. Paired sun and shade comparisons

Most of the cacao sites in the study are in adjacent shade and sun pairs
- can use paired t-tests to control for differences in location
- compare response variables (abundance, richness, diversity)
- note that one shade site was not paired, remove this site from this analysis


##a. create paired_data

```{r}
# retain only paired shade and sun cacao sites
paired_data <- site_data %>%
  filter(Tipo == 'N' | Tipo == 'C',
         !Sitio == 'VC-N')

# site labels for paired visualizations
paired_site_labels <-  c("Sun \n (n = 11)", "Shade \n (n = 11)")
```


##b. abundance

paired t-test
```{r}
# restructure paired_data to wide format for abundance
paired_abundance <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'n')) %>%
  pivot_wider(names_from = Tipo, values_from = n)

# run paired t-test
t.test(paired_abundance$N, paired_abundance$C, paired = TRUE)
```


##c. richness

paired t-test
```{r}
# restructure paired_data to wide format for richness
paired_richness <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'no_species')) %>%
  pivot_wider(names_from = Tipo, values_from = no_species)

# run paired t-test
t.test(paired_richness$N, paired_richness$C, paired = TRUE)
```


##d. diversity indices

Shannon diversity paired t-test
```{r}
# restructure paired_data to wide format for Shannon diversity
paired_shannon <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'Shannon_Index')) %>%
  pivot_wider(names_from = Tipo, values_from = Shannon_Index)

# run paired t-test
t.test(paired_shannon$N, paired_shannon$C, paired = TRUE)
```

Simpson diversity paired t-test
```{r}
# restructure paired_data to wide format for Simpson diversity
paired_simpson <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'Simpson_Index')) %>%
  pivot_wider(names_from = Tipo, values_from = Simpson_Index)

# run paired t-test
t.test(paired_simpson$N, paired_simpson$C, paired = TRUE)
```


##e. make arranged paired figure

paired abundance figure
```{r}
plot_paired_abundance <- ggplot(paired_data, 
                                aes(x = Tipo, y = n, fill = Tipo)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) +
  scale_x_discrete(labels = c(paired_site_labels)) +
  my_theme() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(y = "Abundance") +
  annotate("text", x = 1.5, y = max(paired_data$n, na.rm = TRUE) * .7, 
           label = "ns", size = 8)

plot_paired_abundance
#ggsave("output/plot_paired_abundance.png", width = 10, height = 10)
```

paired richness figure
```{r}
# plot richness between paired sites using palet, my_theme
plot_paired_richness <- ggplot(paired_data, 
                                     aes(x = Tipo, y = no_species, fill = Tipo)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) + 
  scale_x_discrete(labels = c(paired_site_labels)) +
  my_theme() + 
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(y = "Richness") +
  annotate("text", x = 1.5, y = max(paired_data$no_species, na.rm = TRUE) * .85, 
           label = "ns", size = 8)

plot_paired_richness
#ggsave("output/plot_paired_richness.png", width = 10, height = 10)
```

paired simpson figure
```{r}
# plot Simpson Index between paired sites, labels different b/c bottom panel
plot_paired_simpson <- ggplot(paired_data,
                              aes(x = Tipo, y = Simpson_Index, fill = Tipo)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) +
  scale_x_discrete(labels = c(paired_site_labels)) +
  my_theme() +
  theme(axis.title.y = element_text(margin = margin(r = 10)),
        axis.title.x = element_text(margin = margin(t = 10))) +
  labs(x = "Cacao type",
       y = "Simpson\nIndex") +
  annotate("text", x = 1.5, y = max(paired_data$Simpson_Index, na.rm = TRUE) * .7,
           label = "ns", size = 8)

plot_paired_simpson
```

arrange all figures for publication
```{r}
plot_paired_arrange <- ggarrange(
  plot_paired_abundance +
    theme(axis.title.y = element_text(hjust = 0.5),
          plot.margin = margin(5, 5, 5, 20)),
  plot_paired_richness +
    theme(axis.title.y = element_text(hjust = 0.5),
          plot.margin = margin(5, 5, 5, 20)),
  plot_paired_simpson +
    theme(axis.title.y = element_text(hjust = 0.35),
          plot.margin = margin(5, 5, 10, 5)),
  labels = c("a", "b", "c"),
  ncol = 1, nrow = 3,
  align = "v",
  label.x = 0.005,
  label.y = 0.12,
  font.label = list(size = 26, face = "bold"))

plot_paired_arrange
#ggsave("output/plot_paired_arrange.png", width = 10, height = 15)
```


#3. GLMs with habitat type as fixed effect for each response variable

Use GLMs to compare the effect of habitat type on each response variable
- forest and abandoned plantation sites are not paired and have different 
  sample sizes so use GLMs to more robustly compare response variables
- then can run post-hoc test like Tukey's HSD


##a. assess distribution of response variables

assume Poisson for count data
```{r}
hist(site_data$n) # assume Poisson
hist(site_data$no_species) # assume Poisson
hist(site_data$Shannon_Index) # approximately normal
hist(site_data$Simpson_Index) # approximately normal
```


##b. abundance

GLM for abundance
- Poisson model overdispersed
- negative binomial appropriately accounts for the overdispersion
- no significant difference, so no need for post hoc test
```{r}
glm_type_abundance <- glm.nb(n ~ Tipo, 
                       data = site_data)
summary(glm_type_abundance)
```

assess model fit
```{r}
# assess how model fits using DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_type_abundance)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```


##c. richness

GLM for richness
- assume Poisson - it is not overdispersed!
- no significant difference, so no need for post hoc test
```{r}
glm_type_richness <- glm(no_species ~ Tipo,
                         data = site_data,
                         family = poisson)
summary(glm_type_richness)
```

assess model fit
```{r}
# assess how model fits using DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_type_richness)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```


##d. diversity indices

GLM for Shannon diversity
- assume normal - refer to histogram above - not overdispersed!
- significant difference - use Tukey's HSD for post-hoc test
```{r}
glm_type_shannon <- glm(Shannon_Index ~ Tipo,
                         data = site_data)
summary(glm_type_shannon)
```

assess model fit
```{r}
# assess how model fits using DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_type_shannon)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

Tukey's HSD test to detect difference between site types
- accounting for family error rate, no sig differences between site types
```{r}
# Tukey HSD-style pairwise comparisons
tukey_shannon <- glht(glm_type_shannon, linfct = mcp(Tipo = "Tukey"))

# View the summary of pairwise contrasts
summary(tukey_shannon)

# Confidence intervals
confint(tukey_shannon)
```

GLM for Simpson diversity
- assume normal - refer to histogram above - not overdispersed!
- significant difference - use Tukey's HSD for post hoc test
```{r}
glm_type_simpson <- glm(Simpson_Index ~ Tipo,
                         data = site_data)
summary(glm_type_simpson)
```

assess model fit
```{r}
# assess how model fits using DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_type_simpson)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

Tukey's HSD test to detect difference between site types
- accounting for family error rate, no sig differences between site types
```{r}
# Tukey HSD-style pairwise comparisons
tukey_simpson <- glht(glm_type_simpson, linfct = mcp(Tipo = "Tukey"))

# View the summary of pairwise contrasts
summary(tukey_simpson)

# Confidence intervals
confint(tukey_simpson)
```


##e. make arranged GLM  type figure

labels for arranged type figure
```{r}
type_site_labels <-  c("Sun \n (n = 11)", "Shade \n (n = 11)", 
                       "Abandoned \n (n = 4)", "Forest \n (n =5)")
```

abundance across site types figure
```{r}
plot_type_abundance <- ggplot(site_data, 
                                aes(x = Tipo, y = n, fill = Tipo)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) +
  scale_x_discrete(labels = c(type_site_labels)) +
  my_theme() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(y = "Abundance") +
  annotate("text", x = 2.5, y = max(site_data$n, na.rm = TRUE) * .7, 
           label = "ns", size = 8)

plot_type_abundance
#ggsave("output/plot_type_abundance.png", width = 10, height = 10)
```

richness across site types figure
```{r}
plot_type_richness <- ggplot(site_data, 
                                aes(x = Tipo, y = no_species, fill = Tipo)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) +
  scale_x_discrete(labels = c(type_site_labels)) +
  my_theme() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  labs(y = "Richness") +
  annotate("text", x = 2.5, y = max(site_data$no_species, na.rm = TRUE) * .9, 
           label = "ns", size = 8)

plot_type_richness
#ggsave("output/plot_type_richness.png", width = 10, height = 10)
```

Simpson across site types figure
```{r}
# plot Simpson Index across site types, labels different b/c bottom panel
plot_type_simpson <- ggplot(site_data,
                              aes(x = Tipo, y = Simpson_Index, fill = Tipo)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) +
  scale_x_discrete(labels = c(type_site_labels)) +
  my_theme() +
  theme(axis.title.y = element_text(margin = margin(r = 10)),
        axis.title.x = element_text(margin = margin(t = 10))) +
  labs(x = "Land use type",
       y = "Simpson\nIndex") +
  annotate("text", x = 2.5, y = max(paired_data$Simpson_Index, na.rm = TRUE) * .8,
           label = "ns", size = 8)

plot_type_simpson
```

arrange all figures for publication
- maybe just use this figure in the final publication
```{r}
plot_type_arrange <- ggarrange(
  plot_type_abundance +
    theme(axis.title.y = element_text(hjust = 0.5),
          plot.margin = margin(5, 5, 5, 20)),
  plot_type_richness +
    theme(axis.title.y = element_text(hjust = 0.5),
          plot.margin = margin(5, 5, 5, 20)),
  plot_type_simpson +
    theme(axis.title.y = element_text(hjust = 0.35),
          plot.margin = margin(5, 5, 10, 5)),
  labels = c("a", "b", "c"),
  ncol = 1, nrow = 3,
  align = "v",
  label.x = 0.005,
  label.y = 0.12,
  font.label = list(size = 26, face = "bold"))

plot_type_arrange
#ggsave("output/plot_type_arrange.png", width = 10, height = 15)
```


#4. GLMs for each predictor variable for active cacao plantations only

GLMs to assess the influence of local and landscape variables on response vars
- only for active cacao plantations, site type as fixed effect

response vars
- abundance, no_species, Shannon, Simpson, Inv_Simpson

predictor vars
- tipo, scaled: veg, hoja, CC, forest, water, elevation


##a. create active_data

Only assessing impact of local and landscape variables on active plantations
- so filter out abandoned and forest site types
- note that this includes one more site than paired_data
```{r}
active_data <- site_data %>%
  filter(Tipo == 'N' | Tipo == 'C')
```



##b. assess distribution of response variables

histogram of each response var
- will assume Poisson distribution first b/c count data
- assess overdispersion and model fit during model building and adjust as needed
```{r}
hist(active_data$n) # assume Poisson
hist(active_data$no_species) # assume Poisson
hist(active_data$Shannon_Index) # approximately normal
hist(active_data$Simpson_Index) # approximately normal
hist(active_data$Inv_Simpson_Index) # perhaps lognormal?
```


##c. assess correlation of predictor variables

look at correlation matrix - high collinearity if correlation > 0.7
```{r}
ggpairs(active_data, columns = c(18,19,20,21,22,23))
```

canopy cover and hoja are highly correlated (0.873!) - only use canopy cover


##d. plot variable relationships with abundance

```{r}
ggplot(active_data, aes(x = Tipo, y = n)) +
  geom_boxplot()

ggplot(active_data, aes(x = veg_scaled, y = n)) +
  geom_point()

ggplot(active_data, aes(x = CC_scaled, y = n)) +
  geom_point()

ggplot(active_data, aes(x = forest_scaled, y = n)) +
  geom_point()

ggplot(active_data, aes(x = water_scaled, y = n)) +
  geom_point()

ggplot(active_data, aes(x = elevation_scaled, y = n)) +
  geom_point()
```


##e. GLM for abundance with all predictors

Poisson abundance model was overdispersed - use negative binomial instead

full negative binomial glm.nb for abundance
```{r}
glm_full_abundance <- glm.nb(n ~ Tipo + veg_scaled + CC_scaled + forest_scaled + 
                       water_scaled + elevation_scaled, 
                       data = active_data)
summary(glm_full_abundance)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")  

# run dredge to get all possible model combinations - test restricting or not
abundance_model_set <- dredge(glm_full_abundance, m.lim = c(1,3))

# view top models by AIC
head(abundance_model_set)

# get the best model (lowest AICc)
best_abundance_model <- get.models(abundance_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_abundance_model)

# assess multicollinearity
vif(glm_full_abundance)
# vif(best_abundance_model) # contains fewer than two terms

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_abundance_model)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)

# assess how full model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_full_abundance)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

negative binomial abundance model summary

full model:
- combined adjusted quantile test significant, but otherwise no sig deviations
- CC_scaled and Tipo significant

best model:
- no issues with model fit
- contains only CC_scaled - although this is not sig


##f. GLM for richness with all predictors

same process as above except using richness as response variable here

full Poisson glm for richness
```{r}
# full richness model with all predictors
glm_full_richness <- glm(no_species ~ Tipo + veg_scaled + CC_scaled + forest_scaled +
                       water_scaled + elevation_scaled,
                       data = active_data,
                       family = poisson)
summary(glm_full_richness)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")

# run dredge to get all possible model combinations
richness_model_set <- dredge(glm_full_richness, m.lim = c(1,3))

# view top models by AIC
head(richness_model_set)

# get the best model (lowest AICc)
best_richness_model <- get.models(richness_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_richness_model)

# assess multicollinearity
vif(glm_full_richness)
# vif(best_richness_model) model contains fewer than two terms

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_richness_model)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput) # how to interpret this?

# assess how full model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_full_richness)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

Poisson richness model summary

full model:
- no issues with model fit
- nothing sig

best model:
- contains only CC_scaled, not significant
- no issues with model fit


##g. GLM for Shannon diversity with all predictors

full glm for shannon with normal dist.
- Shannon isn't count data! so don't use Poisson? non-integer values?
```{r}
# full shannon model with all predictors
glm_full_shannon <- glm(Shannon_Index ~ Tipo + veg_scaled + CC_scaled + forest_scaled +
                       water_scaled + elevation_scaled,
                       data = active_data,
                       family = gaussian) # normal distribution
summary(glm_full_shannon)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")

# run dredge to get all possible model combinations
shannon_model_set <- dredge(glm_full_shannon, m.lim = c(1,3))

# view top models by AIC
head(shannon_model_set)

# get the best model (lowest AICc)
best_shannon_model <- get.models(shannon_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_shannon_model)

# assess multicollinearity
vif(glm_full_shannon)
vif(best_shannon_model) # inflation not an issue in best model

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_shannon_model)
plot(simulationOutput) # combined adjusted quantile test significant

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)

# assess how full model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_full_shannon)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

normal Shannon diversity model summary

full model:
- no issues with model fit
- nothing sig

best model:
- combined adjusted quantile test significant, no other issues with fit
- contains elevation_scaled and forest_scaled
- elevation sig negative effect
- forest sig positive effect


##h. GLM for Simpson diversity with all predictors

full glm for simpson with normal distribution
- simpson isn't count data! so can't use Poisson distribution
```{r}
# full simpson model with all predictors
glm_full_simpson <- glm(Simpson_Index ~ Tipo + veg_scaled + CC_scaled + forest_scaled +
                       water_scaled + elevation_scaled,
                       data = active_data,
                       family = gaussian)
summary(glm_full_simpson)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")

# run dredge to get all possible model combinations
simpson_model_set <- dredge(glm_full_simpson, m.lim = c(1,3))

# view top models by AIC
head(simpson_model_set)

# get the best model (lowest AICc)
best_simpson_model <- get.models(simpson_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_simpson_model) # elevation and tipo significant effect

# assess multicollinearity
vif(glm_full_simpson)
vif(best_simpson_model) # inflation not an issue in best model

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_simpson_model)
plot(simulationOutput) # combined adjusted quantile test not significant

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)

# assess how full model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_full_simpson)
plot(simulationOutput) 

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

normal Simpson diversity model summary

full model:
- no issues with model fit
- Tipo sig

best model:
- no issues with model fit
- contains both elevation_scaled and Tipo
- elevation has significant negative effect on diversity
- bosque has significantly higher Simpson than all other landuse types








