---
title: "pub_cacao_analysis"
author: "Holden Jones"
date: "2025-07-11"
output: html_document
---

Statistical tests and figures for cacao ranas publication
- using processed dataframes created in pub_cacao_process script


#1. Load

```{r}
library(tidyverse)
library(RColorBrewer)
library(GGally)
library(car)
library(DHARMa)
library(ggResidpanel)
library(ggeffects)
library(MuMIn)
library(MASS) # note that this produces masking issues for select()
```

```{r}
site_data <- read_csv("data/processed_data/site_data_processed.csv")

# reorder factor levels for plotting
site_data$Tipo <- factor(site_data$Tipo,
                         c("N", "C", "V", "B"))
```

define a consistent palet, labels, and theme to be used in figures
```{r}
# palet for comparisons of all 4 site types
pal <- brewer.pal(4, "BrBG")

# set a theme for ggplots
my_theme <- function() {
  theme_minimal() + 
   theme(legend.position = "none",
        plot.background = element_rect("white"),
        panel.background = element_rect("white"),
        panel.grid = element_line("grey90"),
        axis.line = element_line("gray25"),
        axis.text = element_text(size = 16, color = "gray25"),
        axis.title = element_text(size = 22, color = "gray25"),
        legend.text = element_text(size = 16))
}
```


#2. Paired sun and shade comparisons

Most of the cacao sites in the study are in adjacent shade and sun pairs
- can use paired t-tests to control for differences in location
- compare response variables (abundance, richness, diversity)
- note that one shade site was not paired, remove this site from this analysis


##a. create paired_data

```{r}
# retain only paired shade and sun cacao sites
paired_data <- site_data %>%
  filter(Tipo == 'N' | Tipo == 'C',
         !Sitio == 'VC-N')

# site labels for paired visualizations
paired_site_labels <-  c("Shade \n (n = 11)", "Sun \n (n = 11)")
```


##b. abundance

paired t-test
```{r}
# restructure paired_data to wide format for abundance
paired_abundance <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'n')) %>%
  pivot_wider(names_from = Tipo, values_from = n)

# run paired t-test
t.test(paired_abundance$N, paired_abundance$C, paired = TRUE)
```

visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = n, fill = Tipo)) +
  geom_boxplot()
```


## publication level visualization - working

```{r}
# plot species richness between paired sites using palet, my_theme
plot_paired_abundance <- ggplot(paired_data, 
                                     aes(x = Tipo, y = n, fill = Tipo)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) + # order needs to match factor levels above
  scale_x_discrete(labels = c(paired_site_labels)) +
  my_theme() + 
  theme(axis.title.x=element_blank(),
        #axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
  labs(x = "Cacao type",
       y = "Abundance")
plot_paired_abundance
#ggsave("output/plot_paired_abundance.png", width = 10, height = 10)

# below using ggarrange to align multiple figs - don't want to repeat axis labels
# plot_richness_by_site_cacao_arrange <- ggplot(richness_by_site_cacao, 
#                                      aes(x = Tipo, y = no_species, fill = Tipo)) +
#   geom_boxplot() +
#   scale_fill_manual(values = pal) + 
#   scale_y_continuous(limits = c(2, 10)) +
#   scale_x_discrete(labels = c(site_labels)) +
#   my_theme() + 
#   theme(axis.title.x=element_blank(),
#         #axis.text.x=element_blank(),
#         axis.ticks.x=element_blank()) +
#   labs(y = "Richness")
# plot_richness_by_site_cacao_arrange
```












##c. richness

paired t-test
```{r}
# restructure paired_data to wide format for richness
paired_richness <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'no_species')) %>%
  pivot_wider(names_from = Tipo, values_from = no_species)

# run paired t-test
t.test(paired_richness$N, paired_richness$C, paired = TRUE)
```

visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = no_species, fill = Tipo)) +
  geom_boxplot()
```




## publication level visualization - working

```{r}
# plot richness between paired sites using palet, my_theme
plot_paired_richness <- ggplot(paired_data, 
                                     aes(x = Tipo, y = no_species, fill = Tipo)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) + 
  scale_x_discrete(labels = c(paired_site_labels)) +
  my_theme() + 
  labs(x = "Cacao type",
       y = "Richness")
plot_paired_richness
#ggsave("output/plot_paired_abundance.png", width = 10, height = 10)
```






##d. diversity indices

Shannon diversity paired t-test
```{r}
# restructure paired_data to wide format for Shannon diversity
paired_shannon <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'Shannon_Index')) %>%
  pivot_wider(names_from = Tipo, values_from = Shannon_Index)

# run paired t-test
t.test(paired_shannon$N, paired_shannon$C, paired = TRUE)
```

Shannon diversity visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = Shannon_Index, fill = Tipo)) +
  geom_boxplot()
```


## publication level visualization - working

```{r}
# plot Shannon diversity between paired sites using palet, my_theme
plot_paired_shannon <- ggplot(paired_data, 
                                     aes(x = Tipo, y = Shannon_Index, fill = Tipo)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) + 
  scale_x_discrete(labels = c(paired_site_labels)) +
  my_theme() + 
  labs(x = "Cacao type",
       y = "Shannon Index")
plot_paired_shannon
#ggsave("output/plot_paired_shannon.png", width = 10, height = 10)
```


Simpson diversity paired t-test
```{r}
# restructure paired_data to wide format for Simpson diversity
paired_simpson <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'Simpson_Index')) %>%
  pivot_wider(names_from = Tipo, values_from = Simpson_Index)

# run paired t-test
t.test(paired_simpson$N, paired_simpson$C, paired = TRUE)
```

Simpson diversity visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = Simpson_Index, fill = Tipo)) +
  geom_boxplot()
```



```{r}
# plot Simpson diversity between paired sites using palet, my_theme
plot_paired_simpson <- ggplot(paired_data, 
                                     aes(x = Tipo, y = Simpson_Index, fill = Tipo)) +
  geom_boxplot() +
  scale_fill_manual(values = pal) + 
  scale_x_discrete(labels = c(paired_site_labels)) +
  my_theme() + 
  labs(x = "Cacao type",
       y = "Shannon Index")
plot_paired_simpson
#ggsave("output/plot_paired_simpson.png", width = 10, height = 10)
```




## maybe remove inverse simpson here

inverse Simpson diversity paired t-test
```{r}
# restructure paired_data to wide format for inverse Simpson diversity
paired_inv_simpson <- paired_data %>%
  dplyr::select(c('Tipo', 'Pair', 'Inv_Simpson_Index')) %>%
  pivot_wider(names_from = Tipo, values_from = Inv_Simpson_Index)

# run paired t-test
t.test(paired_inv_simpson$N, paired_inv_simpson$C, paired = TRUE)
```

inverse Simpson diversity visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = Inv_Simpson_Index, fill = Tipo)) +
  geom_boxplot()
```



##e. arrange figures for publication








#3. GLMs for each predictor variable

GLMs to assess the influence of local and landscape variables on response vars
- include all site types, site type as fixed effect

response vars
- abundance, no_species, Shannon, Simpson, Inv_Simpson

predictor vars
- tipo, scaled: veg, hoja, CC, forest, water, elevation


##a. assess distribution of response variables

histogram of each response var
- will assume Poisson distribution first b/c count data
- assess overdispersion and model fit during model building and adjust as needed
```{r}
hist(site_data$n) # assume Poisson
hist(site_data$no_species) # assume Poisson
hist(site_data$Shannon_Index) # approximately normal
hist(site_data$Simpson_Index) # approximately normal
hist(site_data$Inv_Simpson_Index) # perhaps lognormal?
```


##b. assess correlation of predictor variables

look at correlation matrix - high collinearity if correlation > 0.7
```{r}
ggpairs(site_data, columns = c(18,19,20,21,22,23))
```

canopy cover and hoja are highly correlated (0.873!) - only use canopy cover


##c. plot variable relationships with abundance

```{r}
ggplot(site_data, aes(x = Tipo, y = n)) +
  geom_boxplot()

ggplot(site_data, aes(x = veg_scaled, y = n)) +
  geom_point()

ggplot(site_data, aes(x = CC_scaled, y = n)) +
  geom_point()

ggplot(site_data, aes(x = forest_scaled, y = n)) +
  geom_point()

ggplot(site_data, aes(x = water_scaled, y = n)) +
  geom_point()

ggplot(site_data, aes(x = elevation_scaled, y = n)) +
  geom_point()
```


## d. GLM for abundance with all predictors

Poisson abundance model was overdispersed - use negative binomial instead

full negative binomial glm.nb for abundance
```{r}
glm_full_abundance <- glm.nb(n ~ Tipo + veg_scaled + CC_scaled + forest_scaled + 
                       water_scaled + elevation_scaled, 
                       data = site_data)
summary(glm_full_abundance)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")  

# run dredge to get all possible model combinations - test restricting or not
abundance_model_set <- dredge(glm_full_abundance, m.lim = c(1,3))

# view top models by AIC
head(abundance_model_set)

# get the best model (lowest AICc)
best_abundance_model <- get.models(abundance_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_abundance_model)

# assess multicollinearity
vif(glm_full_abundance)
# vif(best_abundance_model) # contains fewer than two terms

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_abundance_model)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)

# assess how full model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_full_abundance)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

negative binomial abundance model summary

full model:
- combined adjusted quantile test significant, but otherwise no sig deviations
- CC_scaled and Tipo significant

best model:
- no issues with model fit
- contains only CC_scaled - although this is not sig


##e. GLM for richness with all predictors

same process as above except using richness as response variable here

full Poisson glm for richness
```{r}
# full richness model with all predictors
glm_full_richness <- glm(no_species ~ Tipo + veg_scaled + CC_scaled + forest_scaled +
                       water_scaled + elevation_scaled,
                       data = site_data,
                       family = poisson)
summary(glm_full_richness)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")

# run dredge to get all possible model combinations
richness_model_set <- dredge(glm_full_richness, m.lim = c(1,3))

# view top models by AIC
head(richness_model_set)

# get the best model (lowest AICc)
best_richness_model <- get.models(richness_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_richness_model)

# assess multicollinearity
vif(glm_full_richness)
# vif(best_richness_model) model contains fewer than two terms

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_richness_model)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput) # how to interpret this?

# assess how full model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_full_richness)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

Poisson richness model summary

full model:
- no issues with model fit
- nothing sig

best model:
- contains only CC_scaled, not significant
- no issues with model fit


##f. GLM for Shannon diversity with all predictors

full glm for shannon with normal dist.
- Shannon isn't count data! so don't use Poisson? non-integer values?
```{r}
# full shannon model with all predictors
glm_full_shannon <- glm(Shannon_Index ~ Tipo + veg_scaled + CC_scaled + forest_scaled +
                       water_scaled + elevation_scaled,
                       data = site_data,
                       family = gaussian) # normal distribution
summary(glm_full_shannon)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")

# run dredge to get all possible model combinations
shannon_model_set <- dredge(glm_full_shannon, m.lim = c(1,3))

# view top models by AIC
head(shannon_model_set)

# get the best model (lowest AICc)
best_shannon_model <- get.models(shannon_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_shannon_model)

# assess multicollinearity
vif(glm_full_shannon)
vif(best_shannon_model) # inflation not an issue in best model

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_shannon_model)
plot(simulationOutput) # combined adjusted quantile test significant

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)

# assess how full model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_full_shannon)
plot(simulationOutput)

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

normal Shannon diversity model summary

full model:
- no issues with model fit
- nothing sig

best model:
- combined adjusted quantile test significant, no other issues with fit
- contains elevation_scaled and forest_scaled
- elevation sig negative effect
- forest sig positive effect


##g. GLM for Simpson diversity with all predictors

full glm for simpson with normal distribution
- simpson isn't count data! so can't use Poisson distribution
```{r}
# full simpson model with all predictors
glm_full_simpson <- glm(Simpson_Index ~ Tipo + veg_scaled + CC_scaled + forest_scaled +
                       water_scaled + elevation_scaled,
                       data = site_data,
                       family = gaussian)
summary(glm_full_simpson)
```

use dredge to run model selection - limit to max of three predictors in a model
```{r}
# make sure na.action is set globally for dredge
options(na.action = "na.fail")

# run dredge to get all possible model combinations
simpson_model_set <- dredge(glm_full_simpson, m.lim = c(1,3))

# view top models by AIC
head(simpson_model_set)

# get the best model (lowest AICc)
best_simpson_model <- get.models(simpson_model_set, subset = 1)[[1]]

# summary of the best model
summary(best_simpson_model) # elevation and tipo significant effect

# assess multicollinearity
vif(glm_full_simpson)
vif(best_simpson_model) # inflation not an issue in best model

# assess how best model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = best_simpson_model)
plot(simulationOutput) # combined adjusted quantile test not significant

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)

# assess how full model fits with DHARMa package
simulationOutput <- simulateResiduals(fittedModel = glm_full_simpson)
plot(simulationOutput) 

testDispersion(simulationOutput)
testZeroInflation(simulationOutput)
```

normal Simpson diversity model summary

full model:
- no issues with model fit
- Tipo sig

best model:
- no issues with model fit
- contains both elevation_scaled and Tipo
- elevation has significant negative effect on diversity
- bosque has significantly higher Simpson than all other landuse types








