---
title: "process"
author: "Holden Jones"
date: "2025-07-09"
output: html_document
---

script for data processing

#1. Load libraries

```{r}
library(tidyverse)
library(vegan)
library(labdsv)
```

define file path for processed outputs
```{r}
# this will be consistent for each output in this script
data_dir <- file.path(".", "data/processed_data")
```


#2. new

ranas - individual occurrence data across all sites
```{r}
# read raw ranas data
ranas <- read_csv("data/raw_data/ranas_01_28_25.csv") %>%
  select(-...20, -...21) # remove dummy columns 

# format dates
ranas$sample_date <- as.Date(ranas$Fecha, format = "%d/%m/%y")

ranas <- ranas %>%
  mutate(year = substr(sample_date, 1, 4))

# need to change VC-V to VC-N, turns out this site was not actually abandoned!
ranas <- ranas %>%
  mutate(Sitio = replace(Sitio, Sitio == "VC-V", "VC-N"))

ranas$Tipo[ranas$Sitio == "VC-N"] <- "N"

# change all epibou to epiesp 
# Lopez-Hervas et al. 2024 show that epibou in this area is actually epiesp
ranas$Final_ID[ranas$Final_ID == "epibou"] <- "epiesp"
```

new - only unique ind. from ranas. Used in analyses
```{r}
new <- ranas %>%
  filter(Recap == "N") # excludes recaptured individuals
```


#3. environmental

environmental - environmental data collected for each sampling event
```{r}
# environmental data - change VC-V to VC-N
environmental <- read_csv("data/raw_data/environment_07_12_24.csv")

environmental$sample_date <- as.Date(environmental$Fecha, format = "%d/%m/%y")

environmental <- environmental %>%
  mutate(year = substr(sample_date, 1, 4))

# update VC-N
environmental <- environmental %>%
  mutate(Sitio = replace(Sitio, Sitio == "VC-V", "VC-N"))

environmental$Tipo[environmental$Sitio == "VC-N"] <- "N"
```

'fill in' veg and hoja columns
- these were measured during first site visit in each sampling round
- if no change between subsequent visits, wrote 'same' - vast majority

fill these columns based on values of previous visit
```{r}
environmental <- environmental %>%
  group_by(Sitio, Fecha, Transecto, Bloque) %>%
  mutate(Veg_fill = if_else(Veg == 'same', NA_character_, Veg),
         Hoja_fill = if_else(Hoja =='same', NA_character_, Hoja))

# this needs to also group by year
environmental <- environmental %>%
  group_by(Sitio, year, Transecto, Bloque) %>%
  fill(Veg_fill, .direction = "down") %>%
  fill(Hoja_fill, .direction = "down") %>%
  ungroup()

# now make 'Veg_fill' the only veg column to avoid confusion, same for Hoja
environmental <- environmental %>%
  select(-c('Veg', 'Hoja'))
```

now calculate site means for veg, hoja
```{r}
environmental$Veg_fill <- as.numeric(environmental$Veg_fill)
environmental$Hoja_fill <- as.numeric(environmental$Hoja_fill)

environmental <- environmental %>%
  group_by(Sitio) %>%
  mutate(site_mean_hoja = mean(Hoja_fill),
         site_mean_veg = mean(Veg_fill))

# now only need one row per site - cut down to 32 rows


# remove potentially confusing columns
```



```{r}
mutate( # calculate transect mean canopy cover
    tran_mean_CC = mean(c_across(`CC A`:`CC D`), na.rm = TRUE))








environment_fill_by_trans <- environment_fill %>%
  group_by(Sitio, Transecto) %>% # add Fecha if want each transect visit
  summarize(tran_mean_veg = mean(Veg_fill, na.rm = TRUE),
            tran_mean_hoja = mean(Hoja_fill, na.rm = TRUE)
            )
```





#4. site_data

site_data - this df will have all response and predictor vars for each site

trans_data - location, canopy cover, pair, miscellaneous data for each transect
```{r}
trans_data <- read_csv("data/raw_data/trans_data_08_19_24.csv")

# change VC-V to VC-N
trans_data <- trans_data %>%
  mutate(Sitio = replace(Sitio, Sitio == "VC-V", "VC-N"))

trans_data$Tipo[trans_data$Sitio == "VC-N"] <- "N"

# filter out location info for each transect, notes for finding sites
trans_data <- trans_data %>%
  select(-c('Start UTM 1', 'Start UTM 2', 'End UTM 1', 'End UTM 2', 
            'End Latitude', 'End Longitude', 'Notas'))
```

calculate transect mean canopy cover for each transect
```{r}
trans_data <- trans_data %>%
  rowwise() %>%
  mutate( # calculate transect mean canopy cover
    tran_mean_CC = mean(c_across(`CC A`:`CC D`), na.rm = TRUE)) %>%
  select(-c('CC A', 'CC B', 'CC C', 'CC D')) # now can remove CC quadrat cols
```




##bring in transect vars collected during each site visit with environmental df



















now join this dataframe with local_by_trans
- ONLY INCLUDING PLANTATIONS HERE! SO ONLY N, C
```{r}
local_by_trans <- left_join(environment_fill_by_trans, local_by_trans,
                                 by = c("Sitio", "Transecto"))

# local_by_trans <- local_by_trans %>%
#   select(Sitio, Tipo, Transecto, 
#          tran_mean_veg, tran_mean_hoja, Pair,
#          tran_mean_CC, tran_coef_var_CC,
#          forest_cover_400, water_dist)

# bin all transect measurements to categories used in Islas studies

# Define the breaks (bins) and labels
breaks <- c(0, 1, 5, 10, 25, 50, 75, 95, 100)
labels <- c("0", "1-5", "5-10", "10-25", "25-50", "50-75", "75-95", "95-100")

local_by_trans$veg_bins <- cut(local_by_trans$tran_mean_veg, 
                                   breaks = breaks, 
                                   labels = labels, 
                                   include.lowest = TRUE, 
                                   right = FALSE)

local_by_trans$hoja_bins <- cut(local_by_trans$tran_mean_hoja, 
                                   breaks = breaks, 
                                   labels = labels, 
                                   include.lowest = TRUE, 
                                   right = FALSE)

local_by_trans$CC_bins <- cut(local_by_trans$tran_mean_CC, 
                                   breaks = breaks, 
                                   labels = labels, 
                                   include.lowest = TRUE, 
                                   right = FALSE)
```




standardize all variables for local_by_trans for multivariate models
```{r}
local_by_trans <- local_by_trans %>%
  ungroup() %>% # Ungrouping to scale variables
  mutate(tran_mean_veg_scaled = as.numeric(scale(tran_mean_veg)),
         tran_mean_hoja_scaled = as.numeric(scale(tran_mean_hoja)),
         tran_mean_CC_scaled = as.numeric(scale(tran_mean_CC)),
         forest_cover_400_scaled = as.numeric(scale(forest_cover_400)),
         water_dist_scaled = as.numeric(scale(water_dist)))

# filename <- "local_by_trans_processed.csv"
# filepath <- file.path(data_dir, filename)
# write_csv(local_by_trans, file = filepath)
```


## NEED TO SCALE ELEVATION VAR TOO!!!!!!!!! URGENT




# local_by_site
easier to work at site level - removes lots of strange use cases
- use local_by_trans df from above to calculate site-level local variables 
```{r}
# # format dates, create year column
# local_by_trans$sample_date <- as.Date(local_by_trans$Fecha, format = "%d/%m/%y")
# local_by_trans <- local_by_trans %>%
#   mutate(year = substr(sample_date, 1, 4))

# calculate site-level averages for local vars
local_by_site <- local_by_trans %>%
  group_by(Sitio) %>%
  mutate(site_mean_veg = mean(tran_mean_veg),
         site_mean_hoja = mean(tran_mean_hoja),
         site_mean_CC = mean(tran_mean_CC)
  )


## TEST INCLUDE ELEVATION AS SCALED VAR! DUH

# scale all vars
local_by_site <- local_by_site %>%
  ungroup() %>% # Ungrouping to scale variables
  mutate(site_mean_veg_scaled = as.numeric(scale(site_mean_veg)),
         site_mean_hoja_scaled = as.numeric(scale(site_mean_hoja)),
         site_mean_CC_scaled = as.numeric(scale(site_mean_CC)),
         forest_cover_400_scaled = as.numeric(scale(forest_cover_400)),
         water_dist_scaled = as.numeric(scale(water_dist)))
```









```{r}
site_data <- trans_data %>% 
  filter(Transecto == 1) # just one row (transect) per site

# reorder factor levels now for downstream plotting
site_type$Tipo <- factor(site_type$Tipo, c("N", "C", "V", "B"))

# make Pair a factor for glms
# MAY NEED TO SET PAIR AS FACTOR
#site_type$Pair <- factor(site_type$Pair)


# # output
# filename <- "site_type_processed.csv"
# filepath <- file.path(data_dir, filename)
# write_csv(site_type, file = filepath)
```























#4. cacao_paired

- remove VC-V b/c not paired

df for first analysis goal. compare abundance, richness, diversity b/ween pairs


## combine richness, diversity, abundance_by_site dfs into one df for easier


```{r}
richness_by_site <- new %>%
  group_by(Sitio) %>%
  summarize(species = list(sort(unique(`Final_ID`))),
            no_species = n_distinct(`Final_ID`))
```



just number of ind
```{r}
# note this is same as site_abundance above
abundance_by_site <- new %>%
  group_by(Sitio) %>%
  count()

# join with site_type to carry over tipo, elevation, harvested info
abundance_by_site <- abundance_by_site %>%
  left_join(site_type, by = "Sitio")

abundance_by_site$Tipo <- factor(abundance_by_site$Tipo, c("N", "C", "V", "B"))

# # output
# filename <- "abundance_by_site_processed.csv"
# filepath <- file.path(data_dir, filename)
# write_csv(abundance_by_site, file = filepath)
```


make new_mat - intermediate step for diversity indices from vegan
```{r}
three_column <- new %>%
  group_by(Sitio) %>%
  count(`Final_ID`)

# needs to be converted to data.frame!
three_column <- data.frame(three_column)

new_mat <- matrify(three_column)

# Define the desired order of rows: Shade, Sun, Abandoned, Forest
new_order <- c(1,3,5,8,19,20,22,25,27,29,30,32, # shade (n=12)
               2,4,7,16,18,21,23,24,26,28,31, # sun (n=11)
               9,13,14,17, # abandoned (n=4)
               6,10,11,12,15 # forest (n=5)
               )

# Reorder rows
new_mat <- new_mat[new_order, ]
```



diversity_by_site - each major diversity metric
```{r}
# use diversity function from vegan package
diversity_by_site <- site_type %>%
    mutate(Shannon_Index = diversity(new_mat, index = "shannon")) %>%
    mutate(Simpson_Index = diversity(new_mat, index = "simpson")) %>%
    mutate(Inv_Simpson_Index = diversity(new_mat, index = "invsimpson"))

# # output
# filename <- "diversity_by_site_processed.csv"
# filepath <- file.path(data_dir, filename)
# write_csv(diversity_by_site, file = filepath)
```








