---
title: "pub_cacao_analysis"
author: "Holden Jones"
date: "2025-07-11"
output: html_document
---

Statistical tests and figures for cacao ranas publication
- using processed dataframes created in pub_cacao_process script


#1. Load

```{r}
library(tidyverse)
library(GGally)
library(car)
library(DHARMa)
library(ggResidpanel)
library(ggeffects)
```

```{r}
site_data <- read_csv("data/processed_data/site_data_processed.csv")
```


#2. Paired sun and shade comparisons

Most of the cacao sites in the study are in adjacent shade and sun pairs
- can use paired t-tests to control for differences in location
- compare response variables (abundance, richness, diversity)
- note that one shade site was not paired, remove this site from this analysis


##a. create paired_data

```{r}
# retain only paired shade and sun cacao sites
paired_data <- site_data %>%
  filter(Tipo == 'N' | Tipo == 'C',
         !Sitio == 'VC-N')
```


##b. abundance

paired t-test
```{r}
# restructure paired_data to wide format for abundance
paired_abundance <- paired_data %>%
  select(c('Tipo', 'Pair', 'n')) %>%
  pivot_wider(names_from = Tipo, values_from = n)

# run paired t-test
t.test(paired_abundance$N, paired_abundance$C, paired = TRUE)
```

visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = n, fill = Tipo)) +
  geom_boxplot()
```


##c. richness

paired t-test
```{r}
# restructure paired_data to wide format for richness
paired_richness <- paired_data %>%
  select(c('Tipo', 'Pair', 'no_species')) %>%
  pivot_wider(names_from = Tipo, values_from = no_species)

# run paired t-test
t.test(paired_richness$N, paired_richness$C, paired = TRUE)
```

visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = no_species, fill = Tipo)) +
  geom_boxplot()
```


##d. diversity indices

Shannon diversity paired t-test
```{r}
# restructure paired_data to wide format for Shannon diversity
paired_shannon <- paired_data %>%
  select(c('Tipo', 'Pair', 'Shannon_Index')) %>%
  pivot_wider(names_from = Tipo, values_from = Shannon_Index)

# run paired t-test
t.test(paired_shannon$N, paired_shannon$C, paired = TRUE)
```

Shannon diversity visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = Shannon_Index, fill = Tipo)) +
  geom_boxplot()
```

Simpson diversity paired t-test
```{r}
# restructure paired_data to wide format for Simpson diversity
paired_simpson <- paired_data %>%
  select(c('Tipo', 'Pair', 'Simpson_Index')) %>%
  pivot_wider(names_from = Tipo, values_from = Simpson_Index)

# run paired t-test
t.test(paired_simpson$N, paired_simpson$C, paired = TRUE)
```

Simpson diversity visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = Simpson_Index, fill = Tipo)) +
  geom_boxplot()
```

inverse Simpson diversity paired t-test
```{r}
# restructure paired_data to wide format for inverse Simpson diversity
paired_inv_simpson <- paired_data %>%
  select(c('Tipo', 'Pair', 'Inv_Simpson_Index')) %>%
  pivot_wider(names_from = Tipo, values_from = Inv_Simpson_Index)

# run paired t-test
t.test(paired_inv_simpson$N, paired_inv_simpson$C, paired = TRUE)
```

Shannon diversity visualization
```{r}
ggplot(paired_data, aes(x = Tipo, y = Inv_Simpson_Index, fill = Tipo)) +
  geom_boxplot()
```


#3. GLMs for each predictor variable

GLMs to assess the influence of local and landscape variables on response vars
- include all site types, site type as fixed effect

response vars
- abundance, no_species, Shannon, Simpson, Inv_Simpson

predictor vars
- tipo, scaled: veg, hoja, CC, forest, water, elevation


##a. assess distribution of response variables

histogram of each response var
- will assume Poisson distribution first b/c count data
- assess overdispersion and model fit during model building and adjust as needed
```{r}
hist(site_data$n) # assume Poisson
hist(site_data$no_species) # assume Poisson
hist(site_data$Shannon_Index) # approximately normal
hist(site_data$Simpson_Index) # approximately normal
hist(site_data$Inv_Simpson_Index) # assume Poisson
```


##b. assess correlation of predictor variables

look at correlation matrix - high collinearity if correlation > 0.7
```{r}
ggpairs(site_data, columns = c(18,19,20,21,22,23))
```

canopy cover and hoja are highly correlated (0.873!) - only use canopy cover


##c. GLM for abundance with all predictors

assume Poisson distribution, don't include hoja_scaled due to colinearity

predictor vars: tipo, scaled: veg, hoja, CC, forest, water, elevation
```{r}
glm_abundance <- glm(n ~ Tipo + veg_scaled + 
                       CC_scaled + forest_scaled + 
                       water_scaled + elevation_scaled, 
                   data = site_data,
                   family = poisson)
summary(glm_abundance)
```

likelihood ratio tests
```{r}
Anova(glm_abundance)
```

look at residual, effects plots
```{r}
sim_res_model <- simulateResiduals(fittedModel = glm_abundance)
plot(sim_res_model)

resid_panel(glm_abundance, plots = 'resid', smoother = TRUE)

plot(ggpredict(glm_abundance, terms = c("Tipo", "elevation_scaled")))
```










